---
title: "tidyr & dplyr"
author: "Markus Kainu"
date: "14.01.2015"
layout: default
---

# tidyr & dplyr examples

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(list(echo=TRUE,eval=TRUE,cache=FALSE,warning=FALSE,message=FALSE,fig.dir= "figure/"))
```


- [tidyr & dplyr cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/01/data-wrangling-cheatsheet.pd)
- [Garrett Grolemund - Data Wrangling with R slaidit](https://dl.dropboxusercontent.com/u/5896466/wrangling-webinar.pdf)

# tidyr

##  Create the data sets
```{r}
library(tidyr)
library(dplyr)
# cases
df <- data.frame(country = c("FR", "DE", "US", "FR", "DE", "US", "FR", "DE", "US"),
                 year = c(2011,2011,2011,2012,2012,2012,2013,2013,2013),
                 n = c(7000,5800,15000,6900,6000,14000,7000,6200,13000), stringsAsFactors = FALSE)
cases <- spread(df, year, n)
#                      
df <- data.frame(city = c("New York", "New York", "London", "London", "Beijing", "Beijing"),
                 size = c("large", "small", "large", "small", "large", "small"),
                 amount = c(23,14,22,16,121,56), stringsAsFactors = FALSE)
pollution <- df
# storms
storms <- data.frame(storm = c("Alberto", "Alex", "Allison", "Ana", "Arlene", "Arthur"),
                     wind = c(110,45,65,40,50,45),
                     pressure = c(1007,1009,1005,1013,1010,1010),
                     date = as.Date(c("2000-08-03", "1998-07-27", "1995-06-03", "1997-06-30", "1999-06-11", "1996-06-17")), stringsAsFactors = FALSE)
# songs
songs <- data.frame(song = c("Across the Universe", "Come Together", "Hello, Goodbye", "Peggy Sue"),
                    name = c("John", "John", "Paul", "Buddy"), stringsAsFactors = FALSE)
# artists
artists <- data.frame(name = c("George", "John", "Paul", "Ringo"),
                    plays = c("sitar", "guitar", "bass", "drums"), stringsAsFactors = FALSE)
```


```{r}
cases

gather(cases, # data
       "year", # name of the key variable
       "n", # name of valut var
       2:4) # variables NOT tidy

```


```{r}
pollution

spread(pollution, # data
       size, # class-var
       amount) # amount

```


```{r}
storms

storms2 <- separate(storms, date, c("year", "month", "day"), sep = "-")
storms2

unite(storms2, "date", year, month, day, sep = "-")
```



# dplyr


```{r}
library(dplyr)
library(ggplot2)
tbl_df(diamonds)

diamonds$x %>%
  mean() %>%
  round(2)
```

## select() - subset rows

```{r, eval=FALSE}
storms
# drop vars
select(storms, -storm)
# subset rows
filter(storms, wind >= 50)
# subset rows - multiple conditions
filter(storms,
       wind >= 50,
       storm %in% c("Alberto", "Alex", "Allison"))

```


### More select-commands

```{r, eval=FALSE}
- # Select everything but
: # Select range
contains() # Select columns whose name contains a character string
ends_with() # Select columns whose name ends with a string
everything() # Select every column
matches() # Select columns whose name matches a regular expression
num_range() # Select columns named x1, x2, x3, x4, x5
one_of() # Select columns whose names are in a group of names
starts_with() # Select columns whose name starts with a character string
```


## mutate () - create new vars

```{r}
mutate(storms, ratio = pressure / wind)

mutate(storms,ratio = pressure / wind,
       inverse = ratio^-1)
```

### More mutate fuctions

>All take a vector of values and return a vector of values

```{r, eval=FALSE}
pmin(), pmax() # Element-wise min and max
cummin(), cummax() # Cumulative min and max
cumsum(), cumprod() # Cumulative sum and product
between() # Are values between a and b?
cume_dist() # Cumulative distribution of values
cumall(), cumany() # Cumulative all and any
cummean() # Cumulative mean
lead(), lag() # Copy with values one position
ntile() #Bin vector into n buckets
dense_rank(), min_rank(), percent_rank(), row_number() # Various ranking methods
```


## summarise() - Change unit

```{r}
pollution %>% 
  summarise(median = median(amount), 
            variance = var(amount))

```

### Good summary functions

> All take a vector of values and return a single value

```{r, eval=FALSE}
min(), max() # Minimum and maximum values
mean() # Mean value
median() # Median value
sum() # Sum of values
var, sd() # Variance and standard deviation of a vector
first() # First value in a vector
last() # Last value in a vector
nth() # Nth value in a vector
n() # The number of values in a vector
n_distinct() # The number of distinct values in a vector
```



## Grouped analysis

```{r}
h <- pollution %>% group_by(city)
h
ungroup(h)
```


```{r}
pollution
pollution  %>% group_by(city) %>%
  summarise(mean = mean(amount),
            sum = sum(amount),
            n = n())
```


```{r}
pollution %>% 
  group_by(city) %>% 
  summarise(mean = mean(amount), sum = sum(amount), n = n())
```


```{r, eval=FALSE}
# data(tb)
# tb %>% 
#   group_by(country, year) %>% 
#   summarise(cases = sum(cases)) %>% 
#   summarise(cases = sum(cases))

Error in FUN(X[[1L]], ...) : 
  only defined on a data frame with all numeric variable
```

## Järjestä data

```{r}
arrange(storms, wind)
arrange(storms, -wind)
arrange(storms, desc(wind), desc(date))

```

## the pipe operatos

```{r}

storms %>% 
  filter(wind >= 50) %>% 
  select(storm, pressure)

pollution  %>% group_by(city) %>%
  summarise(mean = mean(amount),
            sum = sum(amount),
            n = n())

storms %>% 
  mutate(ratio = pressure / wind) %>%
  select(storm, ratio)


pollution  %>% group_by(city) %>%
  mutate(mean = mean(amount),
            sum = sum(amount),
            n = n())


```

# join() - merging data.frames

```{r}
library(dplyr)

songs
artists

left_join(songs, artists, by = "name")

inner_join(songs, artists, by = "name")
#outer_join(songs, artists, by = "name")

semi_join(songs, artists, by = "name")
anti_join(songs, artists, by = "name")
```


```{r, eval=FALSE}
dplyr::bind_cols(x,y)
dplyr::bind_rows
dplyr::union
dplyr::intersect
dplyr::setdif
```